A formal treatment of bidirectional typing

### 2024-01-12 tex/ESOP24/S1-intro.tex
[ tex ] Minor changes

While type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types.
For example, the type of a variable can be looked up in the context.
Bidirectional type synthesis combines type \emph{inference} on this subset of terms with type \emph{checking} (based on a given type) on the rest.
Formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
\begin{inlineenum}
  \item $\Gamma |- \isTerm{t} \syn A$ for \emph{synthesis} and 
  \item $\Gamma |- \isTerm{t} \chk A$ for \emph{checking}.
\end{inlineenum}
The former indicates that the type~$A$ is an output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
The algorithm of a bidirectional type synthesiser can often be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
%if a condition called \emph{mode-correctness} is satisfied.
%every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
%As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a term, and does not require unification as Damas--Milner type inference does.
%Moreover, introducing long-distance unification constraints would even undermine the essence of locality in bidirectional type synthesis.

>>>>

While type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types.
For example, the type of a variable can be looked up in the context.
Bidirectional type synthesis combines type \emph{inference} on this subset of terms with type \emph{checking} (based on a given type) on the rest.
Formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
\begin{inlineenum}
  \item $\Gamma |- \isTerm{t} \syn A$ for \emph{synthesis} and 
  \item $\Gamma |- \isTerm{t} \chk A$ for \emph{checking}.
\end{inlineenum}
The former indicates that the type~$A$ is an output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are input.
The algorithm of a bidirectional type synthesiser can often be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
%if a condition called \emph{mode-correctness} is satisfied.
%every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
%As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a term, and does not require unification as Damas--Milner type inference does.
%Moreover, introducing long-distance unification constraints would even undermine the essence of locality in bidirectional type synthesis.

### 2024-01-12 tex/ESOP24/S4-pre-synthesis.tex
[ tex ] Rewrite the example for binding arity with colourful annotations
[ tex ] Minor changes

Erasure of a bidirectional binding signature removes modes and keeps everything else intact; this can be straightforwardly extended by induction to remove modes from a bidirectional typing derivation and arrive at an ordinary typing derivation, which is soundness.
Instead we can remove typing and retain modes, arriving at a mode derivation.
Conversely, if we have both mode and typing derivations for the same term, we can combine them to obtain a bidirectional typing derivation, which is completeness.
In short, soundness and completeness are no more than the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, directed by the same raw term.
All these can be summarised in one theorem (proved by induction).

@@

\begin{figure}
  \centering
  \small
  \begin{tabular}{ r r l }
    & & is in mode~$d$, \\
    $\smash{\boxed{V |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}\,g\,s}}}$
    & $\isTerm{t}$ \hspace{.1em}
    & misses some type annotation iff $g = \BFalse$, and \\
    & & is in mode~$d$ due to an outermost mode cast iff $s = \BFalse$
  \end{tabular}
  \begin{mathpar}
    \inferrule{x \in V}{V |-_{\Sigma, \Omega} \isTerm{x}^{\syn\,\BTrue\,\BTrue}}\,\SynRule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,g\,s}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn\,g\,\BTrue}}\,\SynRule{Anno}
    \and
    \inferrule{V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,g\,\BTrue}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,\BFalse\,\BFalse}}\,\SynRule{Missing}
    \and
    \inferrule{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,g\,\BTrue}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\chk\,g\,\BFalse}}\,\ChkRule{Sub}
    \and
    \inferrule{V, \vec x_1 |-_{\Sigma, \Omega} \isTerm{t_1}^{\dir{d_1}\,g_1\,s_1} \\ \cdots \\ V, \vec x_n |-_{\Sigma, \Omega} \isTerm{t_n}^{\dir{d_n}\,g_n\,s_n}}
    {V |-_{\Sigma, \Omega} \tmOpts^{\dir{d}\,(\bigwedge_i g_i)\,\BTrue}}\,\Rule{Op}
  \end{mathpar}
  \caption{Generalised mode derivations}
  \label{fig:generalised-mode-derivations}
\end{figure}

@@

\begin{proposition}[Annotatability]\label{thm:annotatability}
If\/ $\Gamma |-_{\Sigma,\erase\Omega} t : A$, then there exists~$t'$ such that\/ $t' \sqsupseteq t$ and $\Gamma |-_{\Sigma,\Omega} t' :^\dir{d} A$ for some~$\dir{d}$.
\end{proposition}
Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term of a simply typed language $(\Sigma, \erase\Omega)$ is typable in the ordinary type system, it may not be directly typable in the bidirectional type system $(\Sigma, \Omega)$ due to some missing annotations, but will be if those annotations are added correctly.
In our theory, \cref{thm:annotatability} can be straightforwardly proved by induction on the derivation produced by generalised mode decoration~(\cref{thm:generalised-mode-decoration}) to construct a bidirectional typing derivation in the same mode.
The interesting case is $\SynRule{Missing}$, which is mapped to $\SynRule{Anno}$, adding to the term a type annotation that comes from the given typing derivation.
\begin{figure}
  \centering\small
  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$}
  \begin{mathpar}
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotate A) \sqsupseteq \isTerm{u}}\;\Rule{More}
    \and
    \inferrule{\vphantom{x : \Identifier}}{\isTerm{x} \sqsupseteq \isTerm{x}}\;\Rule{Var}
    \and
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
      {(\isTerm{t} \annotate A) \sqsupseteq (\isTerm{u} \annotate A)}\;\Rule{Anno}
    \and
    \inferrule{\isTerm{t_1} \sqsupseteq \isTerm{u_1} \quad \cdots \quad \isTerm{t_n} \sqsupseteq{u_n}}
      {\tmOpts \sqsupseteq \tmOpus}\;\Rule{Op}
  \end{mathpar}
  \caption{Annotation ordering between raw terms}
  \label{fig:annotation-ordering}
\end{figure}

>>>>

Erasure of a bidirectional binding signature removes modes and keeps everything else intact; this can be straightforwardly extended by induction to remove modes from a bidirectional typing derivation and arrive at an ordinary typing derivation, which is soundness.
Instead we can remove typing and retain modes, arriving at a mode derivation.
Conversely, if we have both mode and typing derivations for the same term, we can combine them to form a bidirectional typing derivation, which is completeness.
In short, soundness and completeness are merely the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, directed by the same raw term.
All these can be summarised in one theorem and proved by induction.

@@

\begin{figure}
  \centering
  \small
  \begin{tabular}{ r r l }
    & & is in mode~$d$, \\
    $\smash{\boxed{V |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}\,g\,s}}}$
    & $\isTerm{t}$ \hspace{.1em}
    & misses some type annotation iff $g = \BFalse$, and \\
    & & is in mode~$d$ due to an outermost mode cast iff $s = \BFalse$
  \end{tabular}
  \begin{mathpar}
    \inferrule{x \in V}{V |-_{\Sigma, \Omega} \isTerm{x}^{\syn\,\BTrue\,\BTrue}}\,\SynRule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,g\,s}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotatecolon A})^{\syn\,g\,\BTrue}}\,\SynRule{Anno}
    \and
    \inferrule{V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,g\,\BTrue}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,\BFalse\,\BFalse}}\,\SynRule{Missing}
    \and
    \inferrule{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,g\,\BTrue}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\chk\,g\,\BFalse}}\,\ChkRule{Sub}
    \and
    \inferrule{V, \vec x_1 |-_{\Sigma, \Omega} \isTerm{t_1}^{\dir{d_1}\,g_1\,s_1} \\ \cdots \\ V, \vec x_n |-_{\Sigma, \Omega} \isTerm{t_n}^{\dir{d_n}\,g_n\,s_n}}
    {V |-_{\Sigma, \Omega} \tmOpts^{\dir{d}\,(\bigwedge_i g_i)\,\BTrue}}\,\Rule{Op}
  \end{mathpar}
  \caption{Generalised mode derivations}
  \label{fig:generalised-mode-derivations}
\end{figure}

@@

\begin{proposition}[Annotatability]\label{thm:annotatability}
If\/ $\Gamma |-_{\Sigma,\erase\Omega} t : A$, then there exists~$t'$ such that\/ $t' \sqsupseteq t$ and $\Gamma |-_{\Sigma,\Omega} t' :^\dir{d} A$ for some~$\dir{d}$.
\end{proposition}
Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term of a simply typed language $(\Sigma, \erase\Omega)$ is typable in the ordinary type system, it may not be directly typable in the bidirectional type system $(\Sigma, \Omega)$ due to some missing annotations, but will be if those annotations are added correctly.
In our theory, \cref{thm:annotatability} can be straightforwardly proved by induction on the derivation produced by generalised mode decoration~(\cref{thm:generalised-mode-decoration}) to construct a bidirectional typing derivation in the same mode.
The interesting case is $\SynRule{Missing}$, which is mapped to $\SynRule{Anno}$, adding to the term a type annotation that comes from the given typing derivation.
\begin{figure}
  \centering\small
  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$}
  \begin{mathpar}
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotatecolon A) \sqsupseteq \isTerm{u}}\;\Rule{More}
    \and
    \inferrule{\vphantom{x : \Identifier}}{\isTerm{x} \sqsupseteq \isTerm{x}}\;\Rule{Var}
    \and
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
      {(\isTerm{t} \annotatecolon A) \sqsupseteq (\isTerm{u} \annotatecolon A)}\;\Rule{Anno}
    \and
    \inferrule{\isTerm{t_1} \sqsupseteq \isTerm{u_1} \quad \cdots \quad \isTerm{t_n} \sqsupseteq{u_n}}
      {\tmOpts \sqsupseteq \tmOpus}\;\Rule{Op}
  \end{mathpar}
  \caption{Annotation ordering between raw terms}
  \label{fig:annotation-ordering}
\end{figure}

### 2024-01-12 tex/ESOP24/S5-synthesis.tex
[ tex ] Minor changes

The interesting part of the theorem is the case for the $\Rule{Op}$ rule and it is rather complex.
Here we shall give its insight first instead of jumping into the details.
Recall that a typing derivation for $\tmOpts$ contains a substitution $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$.
The goal of type synthesis is exactly to define such a substitution $\rho$, so we have to start with an `accumulating' substitution: a substitution $\rho_0$ that is partially defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere.
By mode-correctness, the accumulating substitution~$\rho_i$ will be defined on enough synthesised variables so that type synthesis or checking can be performed on $t_{i}$ with the context $\Gamma, \vars{x}_{i} : \simsub{\Delta_{i}}{\rho_{i}}$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
If we visit a synthesising argument $\synbiarg[i+1]$, then we may extend the domain of $\rho_i$ to include the synthesised variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and thereby \emph{extend} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$ with the unifier.
Then, if we go through every $t_i$ successfully, we will have a total substitution $\rho_n$ by mode-correctness and a derivation of $\Gamma, \vars{x}_i : \Delta_i |-_{\Sigma, \Omega} t_i :^{\dir{d_i}} \simsub{A}{\rho_n}$ for each sub-term $t_i$.

\begin{remark}
To make the argument above sound, it is necessary to compare types and solve a unification problem.
Hence, we assume that the set~$\Xi$ of type variables has a decidable equality, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
To simplify our choice, we may confine $\Xi$ to any set within the family of sets $\Fin(n)$ of naturals less than~$n$, given that these sets have a decidable equality and the arity of a type construct is finite.
Indeed, in our formalisation, we adopt $\Fin(n)$ as the set of type variables in the definition of $\Type_{\Sigma}$\ifarxiv (see \Cref{sec:formalisation} for details)\fi.
For the sake of clarity in presentation, though, we use named variables and just assume that $\Xi$ has a decidable equality.}
\end{remark}
We need some auxiliary definitions for the notion of extension to state the unification problem:
\begin{defn}
By an \emph{extension}\/ $\sigma \geq \rho$ of a partial substitution $\rho$ we mean that the domain $\dom(\sigma)$ of $\sigma$ contains the domain of $\rho$ and $\sigma(x) = \rho(x)$ for every\/ $x$ in $\dom(\rho)$.
  By a \emph{minimal extension}\/ $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho} \geq \rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma \geq \rho$ and $P(\sigma)$.
\end{defn}
\begin{lemma}\label{lem:unify}
  For any\/ $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution\/ $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, 
  \begin{enumerate}
    \item either there is a minimal extension\/ $\bar{\rho}$ of\/ $\rho$ such that\/ $\simsub{A}{\bar{\rho}} = B$,
    \item or there is no extension\/ $\sigma$ of\/ $\rho$ such that\/ $\simsub{A}{\sigma} = B$
  \end{enumerate}
\end{lemma}
This lemma can be inferred from the correctness of first-order unification~\citep{McBride2003,McBride2003a}, or be proved directly without unification.
We are now ready for \cref{thm:bidirectional-type-synthesis-checking}:

>>>>

The interesting part of the theorem is the case for the $\Rule{Op}$ rule.
We shall give its insight first instead of jumping into the details.
Recall that a typing derivation for $\tmOpts$ contains a substitution $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$.
The goal of type synthesis for this case is exactly to define such a substitution~$\rho$, and we have to start with an `accumulating' substitution: a substitution $\rho_0$ that is partially defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere.
By mode-correctness, the accumulating substitution~$\rho_i$ will be defined on enough synthesised variables so that type synthesis or checking can be performed on $t_{i}$ with the context $\Gamma, \vars{x}_{i} : \simsub{\Delta_{i}}{\rho_{i}}$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
If we visit a synthesising argument $\synbiarg[i+1]$, then we may extend the domain of $\rho_i$ to $\rho_{i+1}$ with the synthesised variables $\fv(A_{i + 1})$, provided that type synthesis is successful and that the synthesised type can be \emph{unified with $A_{i+ 1}$}.
If we go through every $t_i$ successfully, then we will have a total substitution $\rho_n$ by mode-correctness and a derivation of $\Gamma, \vars{x}_i : \Delta_i |-_{\Sigma, \Omega} t_i :^{\dir{d_i}} \simsub{A}{\rho_n}$ for each sub-term~$t_i$.

\begin{remark}
To make the argument above sound, it is necessary to compare types and solve a unification problem.
Hence, we assume that the set~$\Xi$ of type variables has a decidable equality, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
To simplify our choice, we may confine $\Xi$ to any set within the family of sets $\Fin(n)$ of naturals less than~$n$, given that these sets have a decidable equality and the arity of a type construct is finite.
Indeed, in our formalisation, we adopt $\Fin(n)$ as the set of type variables in the definition of $\Type_{\Sigma}$\ifarxiv (see \Cref{sec:formalisation} for details)\fi.
For the sake of clarity in presentation, though, we just use named variables and assume that $\Xi$ has a decidable equality.}
\end{remark}
We need some auxiliary definitions for the notion of extension to state the unification problem:
\begin{defn}
By an \emph{extension}\/ $\sigma \geq \rho$ of a partial substitution $\rho$ we mean that the domain $\dom(\sigma)$ of $\sigma$ contains the domain of $\rho$ and $\sigma(x) = \rho(x)$ for every\/ $x$ in $\dom(\rho)$.
  By a \emph{minimal extension}\/ $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho} \geq \rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma \geq \rho$ and $P(\sigma)$.
\end{defn}
\begin{lemma}\label{lem:unify}
  For any\/ $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution\/ $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, either
  \begin{enumerate}
    \item there is a minimal extension\/ $\bar{\rho}$ of\/ $\rho$ such that\/ $\simsub{A}{\bar{\rho}} = B$, or
    \item there is no extension\/ $\sigma$ of\/ $\rho$ such that\/ $\simsub{A}{\sigma} = B$
  \end{enumerate}
\end{lemma}
This lemma can be inferred from the correctness of first-order unification~\citep{McBride2003,McBride2003a}, or be proved directly without unification.
We are now ready for \cref{thm:bidirectional-type-synthesis-checking}:

### 2024-01-12 tex/ESOP24/S6-example.tex
[ tex ] Minor changes

On the other hand, similar to a parser generator, only one specification is needed for a type-synthesiser generator from the user to produce a corresponding synthesiser. 
Such a specification can be derived by extending $(\Sigma_{\bto}, \Omega_{\Lambda}^{\Leftrightarrow})$ accordingly for additional types and constructs with mode-correctness proved by applying \cref{lem:decidability-mode-correctness}, so its type synthesiser follows from \cref{cor:trichotomy} directly.
\fi

>>>>

On the other hand, similar to a parser generator, only one specification is needed for a type-synthesiser generator from the user to produce a corresponding type synthesiser. 
Such a specification can be derived by extending $(\Sigma_{\bto}, \Omega_{\Lambda}^{\Leftrightarrow})$ accordingly for additional types and constructs with mode-correctness proved by applying \cref{lem:decidability-mode-correctness}, so its type synthesiser follows from \cref{cor:trichotomy} directly.
\fi

### 2024-01-12 tex/ESOP24/S8-future.tex
[ tex ] Minor changes

Extending the logical framework \Dedukti~\cite{Assaf2016}, \varcitet{Felicissimo2023}{'s recent \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
They propose a notion of \LF signature with modes, a definition of mode-correctness (called \emph{well-typed} op.\ cit.), and an algorithm of bidirectional type synthesis.
The algorithm is decidable, sound and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and that the set of computational rules is type-preserving, confluent, and strongly normalising.
However, the last assumption is challenging to establish even for specific languages and lacks a general understanding.

@@

We could first focus on simpler cases that the ordinary typing part is still syntax-directed, but each typing rule is allowed to have multiple bidirectional variants, including different orders of their premises.
In such cases, the mode decorator would need to find all mode derivations, but the type synthesiser should still work in a syntax-directed manner on each mode derivation.
Completeness could still take the simple form presented in this paper too.

>>>>

Extending the logical framework \Dedukti~\cite{Assaf2016}, \varcitet{Felicissimo2023}{'s recent \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
They propose a notion of \LF signature with modes, a definition of mode-correctness (called \emph{well-typed} op.\ cit.), and an algorithm of bidirectional type synthesis.
The algorithm is decidable, sound and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and that the set of computational rules is type-preserving, confluent, and strongly normalising.
The last assumption, however, is challenging to establish even for specific languages and lacks a general understanding.

@@

Yet, we could first focus on simpler cases that the ordinary typing part is still syntax-directed, but each typing rule is refined to multiple bidirectional variants, including different orders of its premises.
In such cases, the mode decorator would need to find all mode derivations, but the type synthesiser should still work in a syntax-directed manner on each mode derivation.
Completeness could still take the simple form presented in this paper too.
